Dissertation Proposal

### 2023-11-09 proposal.tex
future plan

I plan to change the tutor such that it asks students to explain
their prediction. I plan to eyeball the explanations to find new
misconceptions. This piece of data also helps me confirm that
anticipated wrong answers represent the expected misconceptions.

\subsection{Reduce Syntactic Bias}%
\label{s:reduce-syntactic-bias}

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
current results find that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from my instruments, there is still a risk of having biase in student
errors. I plan to adjust the process described in
\Cref{s:find-misconceptions-with-tutor} so that it presents programs
in multiple languages: the SMoL version, a JavaScript translation, and
a Python version.

@@

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for any new misconceptions.

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}). My
current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor that
incorporates all these ideas (\Cref{s:tutor}) and have collected
data from all populations. I plan to explore more literature and
improve the tutor if I find new helpful techniques.

@@

Let's reconsider this program from \Cref{s:rq1-expected-results}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which corresponds
to the \miscon{DefByRef} misconception. If an MCQ only has few options
like this example, students have a great chance (in this case, 50\%)
of just guessing the correct answer or successfully using a process of
elimination. So I have been adding additional options that are not
obviously wrong. The tutor is adding options with an ad hoc rule.
Eventually, I plan to generate options by moving and/or copying
numbers in the current options. In this case, my plan would add
\lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ has the
following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
        current options
  \item the ``Other'' option (which, once chosen, allows students to
        enter arbitrary answers)
\end{itemize}

\paragraph{Penalty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving penalty. Designing
the strategy of assigning penalty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because I do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item An alternative common strategy is to grade by completion.
  However, grading by completion alone encourages students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grades as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The tutor provides opportunities for making case comparisons: each
interpreting task presents a SMoL program and the program's result.
These program-result pairs are concrete cases of the semantics of
SMoL. To incorporate the idea of case comparison, the tutor currently
prompts students to compare these cases and then to summarize the
rules of program behavior at the end of \emph{some} tutorials. I plan
to expand this to \emph{all} tutorials and use a consistent prompt.
The prompt should ask about the behavior of new language constructs
and their interaction with all previously introduced constructs.

@@

Prior research argues (\Cref{s:pedagogic-techniques})that illustrating
how programs runs with notional machines (NMs) is beneficial for
students. I have created a web-based notional machine
\begin{center}
  \href{https://lukuangchen.github.io/stacker-2023}{https://lukuangchen.github.io/stacker-2023}
\end{center}

The notional machine can be configured with URL parameters. For
example,
\href{https://lukuangchen.github.io/stacker-2023/?program=\%28defvar+x+12\%29\%0A\%28defvar+y+x\%29\%0A\%28set\%21+y+0\%29\%0Ax\%0Ay}{this
link} opens the NM with a program embedded.

My plan is to change the tutor so that it provides a
program-specific link after each interpreting task and instructs
students to click the link for more explanation.

@@

The current tutor is not in the final shape yet. In particular, many
of the wrong answers correspond to more than one misconception; some
misconceptions are represented by only one question; the order of
interpreting tasks is not randomized. I plan to tackle the first two
differences by adding and removing programs. Adding randomization
requires me to carefully check the dependencies between current
explanations and other texts in the tutor and resolve those
dependencies, if any.

@@

I therefore turned the programs into a set of quizzes (in the US
sense: namely, a brief test of knowledge) that I call the SMoL
Quizzes. There were three quizzes, ordered by linguistic
complexity. The first consisted of only basic operators and
first-order functions. The second added variable and structure
mutation. The third added \lstinline|lambda| and higher-order functions.

Each quiz is a sequence of MCQs. The options are from the Quizius
data. In addition, there is an ``Other'' option. If chosen, the quiz
gave the user a text box with the caption ``Please specify''. The goal
here was to record any other answers, which in turn might lead to
fresh misconceptions.

@@

Question orders were partially randomized. I wanted students to get
some easy, warm-up questions initially, so those were kept at the
beginning. Similarly, I wanted programs that are syntactically
similar to stay close to each other in the quiz. This is so that, when
students got a second such program, they would not have to look far to
find the first one and confirm that they are indeed (slightly)
different, rather than wonder if they were seeing the same program
again.

Students only received feedback after having completed a whole quiz.
At the end of each quiz, they received both summative feedback
\emph{and} a refutation text that explained every program that
appeared in the quiz. Students were also encouraged to run the
programs, but I have little reason to believe that they did (and
certainly they asked few questions on the class forum about them).

Due to space limitations, I present the entire instrument in
\supplRef{SMoL Quizzes/instrument}. Here I focus on a few programs
where student choices correspond to misconceptions identified earlier,
thereby also showing that the curated programs are still effective. I
use a * to indicate the correct answer (which also matches the
implementation's output).

@@

\subsubsection{Quiz 1: Basic Operators and First-Order Functions}%
\label{s:misconceptions-quiz:fun}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Questions}                       &
    \textbf{Table of Answers}                    \\\hline
    \lstinputlisting{programs/def_twice.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$59\% & \lstinline|2 0|             \\
          *32\%       & \lstinline|Error|           \\
          4\%         & \lstinline|Other: 2 2|      \\
          3\%         & ``Nothing is printed'' \\
          2\%         & \lstinline|0 0|             \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/dyn_scope.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \lstinline|Error| \\
          $^\dag$29\% & \lstinline|42 #t| \\
          6\%         & Other        \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/reassign.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          *53\%       & \lstinline|#(6 5) 5|                 \\
          *20\%       & \lstinline|Error|                    \\
          $^\dag$11\% & \lstinline|#(6 6) 5|                 \\
          $^\dag$11\% & \lstinline|#(6 6) 6|                 \\
          5\%         & ``Nothing is printed'' or Other \\
        \end{tabularx}
    }                                            \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizFun. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that I discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-fun}
\end{table}
\Cref{t:quizzes-data-fun} lists programs in \quizFun{} that I
consider the most interesting. These data confirm the presence of
scope-related misconceptions:
\begin{enumerate}
  \item At least 59\% of students incorrectly believe that a
        variable can be defined twice in one block.
  \item 29\% of students believe that a variable defined in a function
        will be available in the top-level (or global) environment
        after the function is called. That is, these students may have a
        dynamic scope misconception.
  \item 22\% (11\% + 11\%) of students believe variables themselves
        can be passed as arguments and redefined inside the function.
        They disagree on whether the redefinition persists after the
        function call.
\end{enumerate}

\subsubsection{Quiz 2: Adding Variable and Structure Mutation}%
\label{s:misconceptions-quiz:state}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \lstinline|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \lstinline|x=#(x x)| or
          something similar. Both \lstinline|(left x)| and \lstinline|(right x)|
          are \lstinline|x| itself.                       \\
          $^\dag$16\% & \lstinline|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    % \lstinputlisting{programs/mvec_in_mvec.txt} & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *62\%       & \lstinline|#(#(100 2 3 4) #(100 2 3 4))| \\
    %       $^\dag$26\% & \lstinline|#(#(1 2 3 4) #(100 2 3 4))|   \\
    %       6\%         & \texttt{Error}                      \\
    %       4\%         & \lstinline|#(#(1 2 3 4) #(1 2 3 4))|     \\
    %       2\%         & Other                               \\
    %     \end{tabularx}
    % }                                               \\\hline
    % \lstinputlisting{programs/vset_in_let.txt}  & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *59\%       & \lstinline|#(10)|  \\
    %       $^\dag$36\% & \lstinline|#(123)| \\
    %       5\%         & Other         \\
    %     \end{tabularx}
    % }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/setset.txt}       & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & Other: \texttt{5 7}   \\
          $^\dag$27\% & \texttt{6 7}          \\
          11\%        & \texttt{5 5}          \\
          2\%         & Other: \texttt{Error} \\
          1\%         & Other: \texttt{5 6}   \\
        \end{tabularx}
    }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that I discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions:
\begin{itemize}
  \item Up to 50\% of students think vectors are copied rather than
        aliased.
  \item 16\% of students think trying to construct and print a
        self-referring vector would error. (This program is ambiguous:
        \emph{constructing} works in most SMoL languages, but \emph{printing}
        may well cause a problem. These identified ambiguities are
        addressed in \Cref{s:tutor-refine-program-set}.)
  \item 31\% of students think a variable is aliased by a parameter if
        the two variables have the same name. Perhaps interestingly,
        fewer students (27\%) think the variable aliasing would
        happen if the two variables had different names.
\end{itemize}

\subsubsection{Quiz 3: Adding Closures and Higher-Order Functions}%
\label{s:misconceptions-quiz:hof}

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \lstinline|3|     \\
          $^\dag$17\% & \lstinline|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \lstinline|4|            \\
          $^\dag$21\% & \lstinline|3|            \\
          6\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    % \lstinputlisting{programs/fun_and_state_2.txt}    & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *88\%       & \lstinline|4|            \\
    %       $^\dag$10\% & \lstinline|3|            \\
    %       3\%         & Other: \lstinline|Error| \\
    %     \end{tabularx}
    % }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|1 1 2 3 2| \\
          $^\dag$34\% & \lstinline|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \lstinline|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that I discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs after the
addition of closures and higher-order functions. It extends what we
saw with loops in \Cref{s:intro}: that students have misconceptions
about their interaction with mutable variables:
\begin{itemize}
  \item 17\% of students think \texttt{lambda} functions can't refer
        to free variables.
  \item 21\% of the students think mutating a variable defined outside
        a lambda can't possibly change the behavior of the lambda.
  \item Student understanding of the
        \lstinline|let|-over-\lstinline|lambda| pattern is weak.
\end{itemize}

@@

\section{Misconceptions Detected by the SMoL Tutor}%
\label{s:misconceptions-tutor}

I now examine what I learned from the interpreting
questions in terms of program behavior
misconceptions. But first, I explain how I associate incorrect
answers with misconceptions.

@@

\subsection{Misinterpreters}
\label{s:misinterp-intro}

In principle, an expert can identify what misconceptions a particular
incorrect answer might correspond to. In practice, I found this
rather difficult for three reasons. First, with 89 programs, each with
several answers, it's easy to make mistakes. Second, my own expert blind
spot may prevent me from seeing an interpretation that would
lead to an additional association. Finally, and specific to my case,
since I had either curated or written all the programs and answers,
I was likely to miss some associations I had not intended.

To address this problem, I formalized misconceptions as
interpreters. That is, for each misconception I created a
corresponding \emph{misinterpreter}. This is an interpreter for the
SMoL syntax that intentionally has a semantic error corresponding to
that misconception. Put differently, a misinterpreter is a like a
``definitional interpreter for a misconception''.

By running programs through the misinterpreters, I can more uniformly
and rigorously identify \emph{all} the misconceptions associated with
a wrong answer. Furthermore, if I identify a new misconception (or
alter a misinterpreter), it is easy to automatically re-classify all
the answers. By using misinterpreters, I indeed found new
interpretations for existing program-answer pairs.

\subsection{A Catalog of Misconceptions}

I iteratively created
our final catalog of misconceptions
(from the perspective of the data in this paper).
I started
with misinterpreters representing the misconceptions described in
\Cref{s:misconceptions-quiz}. These are misconceptions for which we
have reasonable validation (due to the prose in Quizius), so I call
them \emph{grounded} misconceptions. I then looked at wrong answers
not covered by these but chosen by students, and did my best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we
identify with a $^\ddag$), which need to be validated in the future.\footnote{To keep
  the time spent in each module reasonable, the SMoL Tutor does not
  ask students for rationales for their programs. However, I could
  easily modify it to do so a small number of times per user, to make
  progress towards this need.} I then re-ran the misinterpreters
against the chosen answers. I terminated when all the remaining wrong
answers were either (a) found in very few students (we found a gap
between 23\% and 13\%, and hence took 14\% as the threshold),
(b) difficult for
us to attribute to a misconception, or (c) appeared to me to be
``Lispy'' and hence not of broad interest.

\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
list the final catalog.
For each, I also
present a Tutor question
for which the marked wrong answer can be explained by \emph{only} the
named misconception.
That is, that program-answer pair is a representative example of that
misconception.\footnote{An astute reader may well imagine another
  misinterpretation; but I presumably did not find evidence of it in our
  data. I do welcome other misinterpretations, for which we
  can add more misinterpreters!}

@@

\section{Is The Tutor Effective?}
\label{s:effect-tutor}

@@

Recall that the SMoL Tutor is not only a collection of MCQs: it is
also a \emph{tutor}! So far, I have investigated the value of the
MCQs. Now I examine its tutoring aspect. Concretely, I ask:
\begin{center}

  How effective is the Tutor at correcting each misconception?

\end{center}
To study this, I perform the following analysis per
misconception. Using misinterpreters, I identify those questions
whose wrong answers fit only one misconception (i.e., only one output
matches that produced by that misinterpreter, and it matches only that
misinterpreter). I then examine student performance over time across
those questions. This would let me examine how they do on just that
topic, in isolation, over time.

I present the result of this analysis in two forms. Graphically, we
show plots in \Cref{f:tutor-trend-plots}. Each figure shows the percentage of students who
chose the answer corresponding to that misconception. Ideally, we
would like to see these percentages diminish.

@@

These data broadly suggest that the Tutor is a net positive.
Ultimately, however, what these data really show is a need for
improvement in the Tutor. When designing the Tutor questions, many
more were intended to be representative of single
misconceptions. However, as noted in \Cref{s:misinterp-intro}, it is
easy to be incomplete (or incorrect) in ascribing
misconceptions. Furthermore, as their set grows, it is difficult to
reassess all the problems manually. Once evaluated using
misinterpreters, I found far fewer problems than I would have liked.
Concretely, only 40 of the 71 eligible problems (after removing the
non-SMoL modules) were useful in the above
analysis.

>>>>

I plan to change the tutor such that it collects data to uncover these
unknown misconceptions and to confirm known ones.

@@

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for any new misconceptions. I call these
interpreters \emph{misinterpreters}.

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}).
My current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor that
incorporates all these ideas (\Cref{s:tutor}) and have collected data
from all populations.

@@

Let's reconsider this program from \Cref{s:rq1-expected-results}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which corresponds
to the \miscon{DefByRef} misconception. If an MCQ only has few options
like this example, students have a great chance (in this case, 50\%)
of just guessing the correct answer or successfully using a process of
elimination. So I have been adding additional options that are not
obviously wrong. The tutor is adding many options, but with an ad hoc
rule. I plan to revise the rules of adding extra options.

\paragraph{Penalty to Wrong Choices}\label{s:penalty} To discourage
students from making a random choice, the tutor has been giving
penalty. Designing the strategy of assigning penalty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because I do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item An alternative common strategy is to grade by completion.
  However, grading by completion alone encourages students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grades as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

The tutor provides opportunities for making case comparisons: each
interpreting task presents a SMoL program and the program's result.
These program-result pairs are concrete cases of the semantics of
SMoL. To incorporate the idea of case comparison, the tutor currently
prompts students to compare these cases and then to summarize the
rules of program behavior at the end of \emph{some} tutorials.

@@

Prior research argues (\Cref{s:pedagogic-techniques})that illustrating
how programs runs with notional machines (NMs) is beneficial for
students. I have been giving students opportunities and sometimes
explicitly require them to run programs in a NM visualization tool.

@@

The current tutor is not prepared for this analysis yet. I describe how
I plan to change the tutor in \Cref{s:future-work}.

@@

I therefore turned the programs into a set of quizzes (in the US
sense: namely, a brief test of knowledge) that I call the SMoL
Quizzes.
% There were three quizzes, ordered by linguistic
% complexity. The first consisted of only basic operators and
% first-order functions. The second added variable and structure
% mutation. The third added \lstinline|lambda| and higher-order functions.
Each quiz is a sequence of MCQs. The options are from the Quizius
data. In addition, there is an ``Other'' option. If chosen, the quiz
gave the user a text box with the caption ``Please specify''. The goal
here was to record any other answers, which in turn might lead to
fresh misconceptions.

@@

% Question orders were partially randomized. I wanted students to get
% some easy, warm-up questions initially, so those were kept at the
% beginning. Similarly, I wanted programs that are syntactically
% similar to stay close to each other in the quiz. This is so that, when
% students got a second such program, they would not have to look far to
% find the first one and confirm that they are indeed (slightly)
% different, rather than wonder if they were seeing the same program
% again.

% Students only received feedback after having completed a whole quiz.
% At the end of each quiz, they received both summative feedback
% \emph{and} a refutation text that explained every program that
% appeared in the quiz. Students were also encouraged to run the
% programs, but I have little reason to believe that they did (and
% certainly they asked few questions on the class forum about them).

% Due to space limitations, I present the entire instrument in
% \supplRef{SMoL Quizzes/instrument}. Here I focus on a few programs
% where student choices correspond to misconceptions identified earlier,
% thereby also showing that the curated programs are still effective. I
% use a * to indicate the correct answer (which also matches the
% implementation's output).

@@

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Questions}                       &
    \textbf{Table of Answers}                    \\\hline
    \lstinputlisting{programs/def_twice.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$59\% & \lstinline|2 0|             \\
          *32\%       & \lstinline|Error|           \\
          4\%         & \lstinline|Other: 2 2|      \\
          3\%         & ``Nothing is printed'' \\
          2\%         & \lstinline|0 0|             \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/dyn_scope.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \lstinline|Error| \\
          $^\dag$29\% & \lstinline|42 #t| \\
          6\%         & Other        \\
        \end{tabularx}
    }                                            \\\hline
    % \lstinputlisting{programs/reassign.txt}  & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *53\%       & \lstinline|#(6 5) 5|                 \\
    %       *20\%       & \lstinline|Error|                    \\
    %       $^\dag$11\% & \lstinline|#(6 6) 5|                 \\
    %       $^\dag$11\% & \lstinline|#(6 6) 6|                 \\
    %       5\%         & ``Nothing is printed'' or Other \\
    %     \end{tabularx}
    % }                                            \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizFun. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}.}
  \label{t:quizzes-data-fun}
\end{table}
\Cref{t:quizzes-data-fun} lists programs in \quizFun{} that I
consider the most interesting. These data confirm the presence of
scope-related misconceptions.
% :
% \begin{enumerate}
%   \item At least 59\% of students incorrectly believe that a
%         variable can be defined twice in one block.
%   \item 29\% of students believe that a variable defined in a function
%         will be available in the top-level (or global) environment
%         after the function is called. That is, these students may have a
%         dynamic scope misconception.
%   \item 22\% (11\% + 11\%) of students believe variables themselves
%         can be passed as arguments and redefined inside the function.
%         They disagree on whether the redefinition persists after the
%         function call.
% \end{enumerate}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \lstinline|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \lstinline|x=#(x x)| or
          something similar. Both \lstinline|(left x)| and \lstinline|(right x)|
          are \lstinline|x| itself.                       \\
          $^\dag$16\% & \lstinline|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    % \lstinputlisting{programs/mvec_in_mvec.txt} & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *62\%       & \lstinline|#(#(100 2 3 4) #(100 2 3 4))| \\
    %       $^\dag$26\% & \lstinline|#(#(1 2 3 4) #(100 2 3 4))|   \\
    %       6\%         & \texttt{Error}                      \\
    %       4\%         & \lstinline|#(#(1 2 3 4) #(1 2 3 4))|     \\
    %       2\%         & Other                               \\
    %     \end{tabularx}
    % }                                               \\\hline
    % \lstinputlisting{programs/vset_in_let.txt}  & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *59\%       & \lstinline|#(10)|  \\
    %       $^\dag$36\% & \lstinline|#(123)| \\
    %       5\%         & Other         \\
    %     \end{tabularx}
    % }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    % \lstinputlisting{programs/setset.txt}       & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *59\%       & Other: \texttt{5 7}   \\
    %       $^\dag$27\% & \texttt{6 7}          \\
    %       11\%        & \texttt{5 5}          \\
    %       2\%         & Other: \texttt{Error} \\
    %       1\%         & Other: \texttt{5 6}   \\
    %     \end{tabularx}
    % }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions.
% :
% \begin{itemize}
%   \item Up to 50\% of students think vectors are copied rather than
%         aliased.
%   \item 16\% of students think trying to construct and print a
%         self-referring vector would error. (This program is ambiguous:
%         \emph{constructing} works in most SMoL languages, but \emph{printing}
%         may well cause a problem. These identified ambiguities are
%         addressed in \Cref{s:tutor-refine-program-set}.)
%   \item 31\% of students think a variable is aliased by a parameter if
%         the two variables have the same name. Perhaps interestingly,
%         fewer students (27\%) think the variable aliasing would
%         happen if the two variables had different names.
% \end{itemize}

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \lstinline|3|     \\
          $^\dag$17\% & \lstinline|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \lstinline|4|            \\
          $^\dag$21\% & \lstinline|3|            \\
          6\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    % \lstinputlisting{programs/fun_and_state_2.txt}    & {
    %     \begin{tabularx}{\linewidth}{r X}
    %       *88\%       & \lstinline|4|            \\
    %       $^\dag$10\% & \lstinline|3|            \\
    %       3\%         & Other: \lstinline|Error| \\
    %     \end{tabularx}
    % }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|1 1 2 3 2| \\
          $^\dag$34\% & \lstinline|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \lstinline|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs after the
addition of closures and higher-order functions. It extends what we
saw with loops in \Cref{s:intro}: that students have misconceptions
about their interaction with mutable variables.

@@

\section{Preliminary List of Misconceptions}%
\label{s:misconceptions-tutor}

I now examine what I learned from the tutor's interpreting tasks in
terms of program behavior misconceptions.

@@

% \subsection{Misinterpreters}
% \label{s:misinterp-intro}

% In principle, an expert can identify what misconceptions a particular
% incorrect answer might correspond to. In practice, I found this
% rather difficult for three reasons. First, with 89 programs, each with
% several answers, it's easy to make mistakes. Second, my own expert blind
% spot may prevent me from seeing an interpretation that would
% lead to an additional association. Finally, and specific to my case,
% since I had either curated or written all the programs and answers,
% I was likely to miss some associations I had not intended.

% To address this problem, I formalized misconceptions as
% interpreters. That is, for each misconception I created a
% corresponding \emph{misinterpreter}. This is an interpreter for the
% SMoL syntax that intentionally has a semantic error corresponding to
% that misconception. Put differently, a misinterpreter is a like a
% ``definitional interpreter for a misconception''.

% By running programs through the misinterpreters, I can more uniformly
% and rigorously identify \emph{all} the misconceptions associated with
% a wrong answer. Furthermore, if I identify a new misconception (or
% alter a misinterpreter), it is easy to automatically re-classify all
% the answers. By using misinterpreters, I indeed found new
% interpretations for existing program-answer pairs.

% \subsection{A Catalog of Misconceptions}

I iteratively created our final catalog of misconceptions (from the
perspective of the data in this paper). I started with misinterpreters
representing the misconceptions described in
\Cref{s:misconceptions-quiz}. These are misconceptions for which we
have reasonable validation (due to the prose in Quizius), so I call
them \emph{grounded} misconceptions. I then looked at wrong answers
not covered by these but chosen by students, and did my best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we identify with a $^\ddag$), which need to be
validated in the future. I then re-ran the misinterpreters against the
chosen answers. I terminated when all the remaining wrong answers were
either (a) found in very few students (we found a gap between 23\% and
13\%, and hence took 14\% as the threshold), (b) difficult for us to
attribute to a misconception, or (c) appeared to me to be ``Lispy''
and hence not of broad interest.

\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
list the final catalog. For each, I also present a Tutor question for
which the marked wrong answer can be explained by \emph{only} the
named misconception. That is, that program-answer pair is a
representative example of that misconception.

@@

\section{How Effective Is the Tutor?}
\label{s:effect-tutor}

@@

% Recall that the SMoL Tutor is not only a collection of MCQs: it is
% also a \emph{tutor}! So far, I have investigated the value of the
% MCQs. Now I examine its tutoring aspect. Concretely, I ask:
% \begin{center}

%   How effective is the Tutor at correcting each misconception?

% \end{center}
% To study this, I perform the following analysis per
% misconception. Using misinterpreters, I identify those questions
% whose wrong answers fit only one misconception (i.e., only one output
% matches that produced by that misinterpreter, and it matches only that
% misinterpreter). I then examine student performance over time across
% those questions. This would let me examine how they do on just that
% topic, in isolation, over time.

This section presents part of my preliminary answer to \textbf{RQ2}
(\Cref{s:plan-rq2}): is the tutor effective?

I present the result in two forms. Graphically, we show plots in
\Cref{f:tutor-trend-plots}. Each figure shows the percentage of
students who chose the answer corresponding to that misconception.
Ideally, we would like to see these percentages diminish.

@@

Overall, the data suggest that the Tutor \emph{most likely did not do
harm} and perhaps even \emph{may have done some good}. Revisions on
the tutor (\Cref{s:future-work}), in light of misinterpreters, is
necessarily to give a more accurate evaluation. For example I found
far fewer problems than I would have liked. Concretely, only 40 of the
71 eligible problems (after removing the non-SMoL modules) were useful
in the above analysis.

% These data broadly suggest that the Tutor is a net positive.
% Ultimately, however, what these data really show is a need for
% improvement in the Tutor. When designing the Tutor questions, many
% more were intended to be representative of single misconceptions.
% However, as noted in \Cref{s:misinterp-intro}, it is easy to be
% incomplete (or incorrect) in ascribing misconceptions. Furthermore, as
% their set grows, it is difficult to reassess all the problems
% manually. Once evaluated using misinterpreters,
